---
title: "P. rapae Sperm GO enrichment"
author: "Katie McLaughlin & James R. Walters"
output: 
  html_document: 
    highlight: pygments
    theme: readable
    toc: true
    toc_float: true
    toc_collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here we will do GO-enrichment tests with the `GOstats` package, following the procedure outlined in the vignette for "[Hypergeometric testing with unsupported model organisms](https://www.bioconductor.org/packages/release/bioc/html/GOstats.html)".

# Load libraries & data

```{r}
suppressMessages(library(GOstats))
suppressMessages(library(GSEABase))
library(kableExtra)
```

## Sperm Proteins
```{r}
#DE output from DEP
pieris.DE <- read.csv(file = "../../Part1_Proteomics/Results/pierisSpermProtDiffEx.csv", 
header = TRUE)
```


Partition sperm proteins by differential abundance results. Extract sets of unique GeneIDs for Apy-, Eup-, and Un-biased genes. 

```{r}
UnbiasedProteins <- pieris.DE[pieris.DE$master.de == "unbiased", ]
Eu.DE <- pieris.DE[pieris.DE$master.de == "eupyrene", ]
A.DE <- pieris.DE[pieris.DE$master.de == "apyrene", ]

# UnbiasedProteins <- pieris.DE[pieris.DE$significant_bh == FALSE, ]
allspermprots <- unique(pieris.DE$GeneID)
sharedproteins <- unique(UnbiasedProteins$GeneID)
Eupy.proteins <- unique(Eu.DE$GeneID)
Apy.proteins <- unique(A.DE$GeneID)

```

## PANNZER GO

Read in Pannzer GO results and format for working with `GOstats`
```{r}
prapaeGO <- read.delim("../PANNZEROutputs/GO.out")
GeneIdentifiers <- read.csv("../../Part0_DataCleaning/Results/TrxID.GeneName.GeneID.csv")
prapaeGO.all <- merge(x = prapaeGO, y = GeneIdentifiers[ , c(1,3)], 
                      by.x = "qpid", by.y = "TrxID") #Need geneID for transcript in GO.out
goframeData.prapae <- data.frame("go_id" = sprintf("GO:%07d", prapaeGO.all$goid), 
                                 "Evidence" = "ISS", "gene_id" = prapaeGO.all$GeneID)

allprots.prapae <- unique(prapaeGO.all$qpid) #Prapae sperm proteins with GO annotations

```

Format as `geneSetCollection` object.
```{r}
gsc.prapae <- GeneSetCollection(GOAllFrame(GOFrame(goframeData.prapae, 
                                                   organism = "Pieris rapae")), 
                                setType = GOCollection())

```

# GO Enrichment tests

## Configure parameters

To begin with, make a function that generates the relevant set of test parameters, so we will test each GO Class for each portion of the sperm proteome versus the WHOLE sperm proteome
```{r}
ontologyClasses <- c("CC", "BP", "MF")

makeparams <- function(subsetName, genes) {
  params.cats <- lapply(ontologyClasses, function(x) {
    GSEAGOHyperGParams(
      name = subsetName, 
      geneSetCollection = gsc.prapae, 
      geneIds = genes, 
      universeGeneIds = allspermprots, 
      ontology = x, 
      pvalueCutoff = 0.05, 
      conditional = TRUE,
      testDirection = "over"
      )  
    } 
  )
  names(params.cats) <- paste0(subsetName, "_", ontologyClasses)
  return(params.cats)
}
```

And now make param objects for each subset of the proteome.

```{r}
params.apyrene <- makeparams(subsetName = "apyrene", genes = Apy.proteins) #Parameter object for each of the ontology 
params.eupyrene <- makeparams(subsetName = "eupyrene", genes = Eupy.proteins)  
params.shared <- makeparams(subsetName = "shared", genes = sharedproteins) 
```


## Hypergeom Tests
And then execute the Hypergeometric test for each of these subsets. Save R object out to file. 

```{r}
#params.list <- list(params.apyrene, params.eupyrene, params.shared)

HG.list <- lapply(c(params.apyrene, params.eupyrene, params.shared), FUN = hyperGTest) 
save(HG.list, file = "../Results/HypergeomTest.list.RData")

```

## Write Tables

Here we will iterate over each combinaton of GO-class and sperm subset, writing out each set of results in turn to the same file.

I also found that html reports are available for each GO-result object. So I've written these out at the same time.

```{r}

filename <- "../Results/prapaeGOenrichment_out.txt"
# Organize output by GO type, then by sperm subset:

hg.names <- names(HG.list)

out.order <- c(grep("BP", hg.names), grep("MF", hg.names), grep("CC", hg.names))

cat("Gene Ontology Enrichment using Bioconductor GOstat for P. rapae sperm proteins\n\n",  file = filename)

for (i in out.order) {
  cat("\n\n", hg.names[i],":\n", file = filename, append = T)
  
  suppressWarnings( # suppress "appending" warnings
    write.table(summary(HG.list[[i]]), file = filename, 
              quote = F, sep = "\t", row.names = F, append = T)
  )
  # make HTML reports for each result.
  htmlReport(HG.list[[i]], file = paste0("../Results/GO_html_tables/", hg.names[i], ".html"))
}
```



In this case, the data will output in TIDY fashion. This makes it easier to include as supplement in paper, I believe.

```{r}

filename <- "../Results/prapaeGOenrichment_TIDY.txt"
# Organize output by GO type, then by sperm subset:

GOcats <- c("CC" = "Cellular Component", "BP" = "Biological Process", "MF" = "Molecular Function")

#cat("Gene Ontology Enrichment using Bioconductor GOstat for P. rapae sperm proteins\n\n",  file = filename)
cat("Sperm Bias", "GO category",names(summary(HG.list[[1]])),"\n", sep = "\t", file = filename)

for (i in out.order) {
  outGO <- summary(HG.list[[i]])
  sperm_GOtype <- unlist(strsplit(hg.names[i], split = "_", fixed = T))
  spermtype <- rep(sperm_GOtype[1], length(nrow(outGO)))
  GOtype <- rep(GOcats[sperm_GOtype[2]], length(nrow(outGO)))
  names(GOtype) <- NULL # gets rid of warning from dataframe about rownames
  outGO <- cbind("SpermClass" = spermtype,
                  "GO_Category" = GOtype,
                  outGO)
  
  suppressWarnings( # suppress "appending" warnings
  write.table(outGO, file = filename, col.names = FALSE,
              quote = F, sep = "\t", row.names = F, append = T)
  )
}
```

# GO by Gene

For some particular enriched GO categories, it is desirable to examine exactly which genes are supporting that enrichment.

Let's write a function that will identify the genes associated with any particular set of GO IDs.

```{r}
# INPUTS
# myGO: vector of terms
# goRes: a GOHyperGResult object

getGOgenes <- function(myGO, goRes = HG.list[["apyrene_BP"]]) {
  genesByTerm <- geneIdsByCategory(goRes) # list of genes by GO id
  GOtermsAll <- names(genesByTerm) # vector of all GO enriched id

  # Report warning for non-matched terms
  missingGO <- (! myGO %in% GOtermsAll)
  if (any(missingGO)) {
    warning("The following GO ids are not found:\n", 
            paste(myGO[which(missingGO)], "\t\t\n"))
  }
  
  godf <- reshape2::melt(genesByTerm) # melt list into dataframe
  names(godf) <- c("GeneID", "GOid")
  myGOgenes <- godf[ godf$GOid %in% myGO, ]
  return(myGOgenes)
}

# small test
# getGOgenes(myGO = c("GO:007037", "GO:009015", "GO:000609", "GO:0015698") )


```

And we need to cross reference this with the annotations.  Load `allAnno` dataframe from previous script.

```{r}
load(file = "../Results/PrapaeSpermProts_allAnnot.Rdata")
```

## ERK signaling

One Apyrene-enriched GO:BP term that caught attention in review was `GO:0070372`, "regulation of ERK1 and ERK2 cascade".

```{r}
erkGOgenes <- getGOgenes(myGO = c("GO:0070372"))
kable(erkGOgenes) |> kable_styling()
```

```{r}
erkAnno <- merge(erkGOgenes, allAnno, by  = "GeneID")
kable(erkAnno) |> kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 9) |> scroll_box(width = "100%", height = "200px")
```

## Mitochondrial

It appears that the apyrene-enriched proteins generate enrichment for many mitochondrial-related terms.

### GO-BP Mitochondrial

Several apyrene-enriched GO:BP terms indicate mitochondrial function:

* GO:0090151	establishment of protein localization to mitochondrial membrane
* GO:0006099	tricarboxylic acid cycle
* GO:0007008	outer mitochondrial membrane organization
* GO:0070585	protein localization to mitochondrion
* GO:0007007	inner mitochondrial membrane organization
* GO:0006839	mitochondrial transport
* GO:0007006	mitochondrial membrane organization
* GO:0045333	cellular respiration
* GO:0006626	protein targeting to mitochondrion

So we can extract the relevant genes.  It turns out, many of the same genes are supporting different terms.
```{r}
mitoGOBP <- c( "GO:0090151", "GO:0006099", "GO:0007008", "GO:0070585", "GO:0007007",
             "GO:0007006", "GO:0045333", "GO:0006626")
mitoGOBPgenes <- getGOgenes(myGO = mitoGOBP, goRes = HG.list[["apyrene_BP"]])
kable(mitoGOBPgenes) |> kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 9)
table(mitoGOBPgenes$GeneID)

```


<!-- So if we limit to the `r length(unique(mitoGOBPgenes$GeneID))` genes for examining GO:BP annotations, we get: -->

<!-- ```{r} -->
<!-- mitoBPAnno <- allAnno[ allAnno$GeneID %in% unique(mitoGOBPgenes$GeneID), ] -->
<!-- kable(mitoBPAnno) |> kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 9) -->
<!-- ``` -->



### GO-CC Mitochondrial

Several of the apyrene GO:CC terms are also related to mitochondria:

* GO:0005741	mitochondrial outer membrane
* GO:0005758	mitochondrial intermembrane space
* GO:0098798	mitochondrial protein-containing complex
* GO:0005740	mitochondrial envelope
* GO:0005739	mitochondrion
* GO:0005743	mitochondrial inner membrane
* GO:0042645	mitochondrial nucleoid


So we can extract the relevant genes.  It turns out, many of the same genes are supporting different terms.
```{r}
mitoGOCC <- c( "GO:0005741",	"GO:0005758",	"GO:0098798",	"GO:0005740",
               "GO:0005739",	"GO:0005743",	"GO:0042645")
mitoGOCCgenes <- getGOgenes(myGO = mitoGOCC, goRes = HG.list[["apyrene_CC"]])

table(mitoGOCCgenes$GeneID)
```

How much to these overlap with the GO:BP genes?

```{r}
table(unique(mitoGOCCgenes$GeneID) %in% unique(mitoGOBPgenes$GeneID))
```

There is some overlap, but only by about half! Let's combine these into a single master dataframe.

```{r}
apyGOCC <- data.frame(GeneID = unique(mitoGOCCgenes$GeneID),
                      GOCC = rep(TRUE, length(unique(mitoGOCCgenes$GeneID)))
)
apyGOBP <- data.frame(GeneID = unique(mitoGOBPgenes$GeneID),
                      GOBP = rep(TRUE, length(unique(mitoGOBPgenes$GeneID)))
)

apymitoGOgenes <- merge(apyGOCC, apyGOBP, by = "GeneID", all = T)
apymitoGOgenes$GOCC[is.na(apymitoGOgenes$GOCC)] <- FALSE
apymitoGOgenes$GOBP[is.na(apymitoGOgenes$GOBP)] <- FALSE
table(apymitoGOgenes[,2:3])
                      
```

So we have a total of `r nrow(apymitoGOgenes)` different genes that are apyrene-biased and supporting mitochondria-related GO terms.


### Merge & Extract annotations

Ultimately, we want to look at the functional predictions specifically associated with these `r nrow(apymitoGOgenes)` apyrene-biased genes with mito-related functions.  So merge the geneset with the annotations.

```{r}
apyMitoAnno <- merge(x = apymitoGOgenes, y = allAnno, by = "GeneID")
headersWanted <- c("GeneID", "GeneName", "GOCC", "GOBP", "TrxID", "Apyrene_vs_Eupyrene_p.bh", "significant_bh", "master.de", "desc", "subject_seq.id", "expect_value", "bit_value", "organism_commonName", "organism_scientificName", "organism_group", "function.", "SigP_Pred", "TargP_Pred", "transmem")
apyMitoAnno <- apyMitoAnno[, headersWanted]


#  allAnno[ allAnno$GeneID %in% unique(mitoGOBPgenes$GeneID), ]
kable(apyMitoAnno) |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                font_size = 9) |>
  scroll_box(width = "100%", height = "400px")
```


<!-- ```{r} -->
<!-- kable(mitoGOCCgenes) |> kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 9) -->
<!-- table(mitoGOBPgenes$GeneID) -->

<!-- ``` -->


<!-- So if we limit to the `r length(unique(mitoGOBPgenes$GeneID))` genes for examining annotations, we get: -->

<!-- ```{r} -->
<!-- mitoBPAnno <- allAnno[ allAnno$GeneID %in% unique(mitoGOBPgenes$GeneID), ] -->
<!-- kable(mitoBPAnno) |> kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 9) -->
<!-- ``` -->




## Write GO gene tables

```{r}
write.csv(x = apyMitoAnno, file = "../Results/apyMitoGOgenesAnnotated.csv",
          row.names = F)
```


# Session Info
```{r}
sessionInfo()
```

